Investigating social change during cultural contact period using geometric morphometry of pottery shapes from Iron Age northeast Taiwan

### 2019-11-11 analysis/paper/paper.Rmd
updates for backgrounds of GM

Geometric morphometry methods (GMM) has been widely applied recently in archaeology for shape analysis, which explores morphological variability and similarity of archaeological materials to address the questions of anthropological interests. Different from linear measurement approach that captures shapes by measuring distances, angles or ratios of objects, geometric morphometrics methods use Cartesian coordinates of morphological structures to define the shapes. Landmarks, curves or outlines of objects can be transformed into coordinates by their unique location and orientation with respect to numerical values of coordinate axes [@Adams2004; @Lawing2010; @Slice2007]. According to different focuses and forms of original shape data, there are two common morphometric methods used to capture and analyze the shapes, landmark analysis and outline analysis [@Adams2004]. Landmark approaches assign a set of landmark points to specimen that can be specified on a coordinate system. The landmark coordinates are represented by numerical values with respect to defined coordinate axes. Based on the specific position of each point, objects are translated, rotated, and scaled into the same orientation and size using Procrustes superimposition technique to find the best fit between configuration shapes [@Lawing2010]. Landmarks

@@

# Hypotheses and expectations
The goal for locally made pottery data collection is to explore whether pottery production could reflect the social change, from more corporate modes to more network modes after European contact. The level of craft specialization usually relates to the presence of elite control. In this case, if competitions for prestige and European resources among individuals gradually lead to change in social organization, then the production of pottery will show more homogeneous features after European contact due to the craft specialization. The standardization of metric attributes for pottery will be used to measure the craft specialization.

>>>>

Geometric morphometry methods (GMM) has been widely applied recently in archaeology for shape analysis, which explores morphological variability and similarity of archaeological materials to address the questions of anthropological interests. Different from linear measurement approaches that capture shapes by measuring distances, angles or ratios of objects, geometric morphometrics methods use Cartesian coordinates of morphological structures to define the shapes [@Adams2004; @Bookstein1997; @Lawing2010; @Slice2007]. Landmarks, curves or outlines of objects can be represented by coordinates with their unique location and orientation with respect to numerical values of coordinate axes. According to different focuses and forms of original shape data, there are two common morphometric methods used to capture and analyse the shapes, landmark approaches and outline approaches [@Adams2004]. Landmark approaches assign a set of landmarks onto specimens as reference points that can be specified on a coordinate system as x, y coordinates for two dimensions or x, y, z coordinates for three dimensions. Based on the distances and angles of landmark points, the objects are translated, rotated, and scaled into the same orientation and size using generalized Procrustes analysis (GPA) [@Bookstein1991]. GPA method extracts shape information and compares shapes by taking sets of landmark configurations into a common coordinate system, where coordinate values represent shape variables that can be further analysed by further multivariate statistical analyses [@Slice2007]. Landmark approach is a useful method for archaeological objects with extreme morphological signature or distinctive curvature that provide suitable locations for landmark placement, such as projectile point tips or biologically meaningful points indicating osteological differences [@Birch2019; @Buchanan2019; @Lycett2013; @Cardillo2010; @Haruda2019; @Meloro2015].

However, landmark approaches have a limitation that a set of landmarks may not be able to capture the shape differences of a morphological structure when the curve between landmarks conveys more important information than landmarks themselves. In addition, suitable locations for landmark placement may not be available for some complex or rounded shaped structures, such as pottery.

@@

# Hypotheses and expectations

My hypothesis is that if foreign contacts including European and Chinese had impact on the emergence of social inequality in local indigenous society due to the monopolies for trade between a small number of indigenous people, then I expect to see the changes in social organization from a more corporate to a more network organization that can be reflected by pottery production at Kiwulan. The emergence of craft specialization, here pottery specialization, are usually related to the presence of elite control. In this case, if the competitions for foreign resources and being trade partners of European or Chinese colonizer among individuals gradually lead to the emergence of social inequality, then I expect the local pottery will show more homogeneous features after contact due to the pottery specialization caused by control of small group of individuals.
